Machine Learning for Predictive Auto-Tuning with Boosted Regression Trees

### 2012-03-02 paper/final.tex
Main figure fixes
Fix fig5
Add some more future work

The paper is organized as follows:
Section~\ref{sec:brtree} describes the boosted regression tree model and the procedure for fitting it to empirical timing data.
Section~\ref{sec:kernel} describes the sort of kernel we employ for our benchmarking,
Section~\ref{sec:exp} presents the results of our benchmarking experiments, which
compare a reference implementation to a) empirical auto-tuning over a domain-specific grid,
b) empirical auto-tuning over a hill-climbing search, and
c) predictive auto-tuning.
Section~\ref{sec:discuss} provides discussion of our results, and
Section~\ref{sec:fw} outlines directions for future work.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Predictive Auto-Tuning}

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Filterbank Correlation}

\label{sec:fbcorr}

@@

The high potential for arithmetic intensity makes the GPU an ideal platform for computing filterbank correlations,
and and filterbank correlation is used extensively in image and video processing,
where it is often a computational bottleneck.
One might expect then, that it would be easy to implement a library providing
this operation as a simple function that takes pointers and strides for $x$, $f$, and $z$ and performs the computation.
However, as shown in \citet{pinto+cox:2011gcg} 
it is challenging to provide an implementation
or even an implementation strategy that provides satisfactory performance
across the range of inputs (shapes, physical layouts) that occur in
typical usage.
\citet{kamil+etal:2009} summarize a related situation related to general stencil computations in their abstract:
``Although the auto-tuning strategy has been successfully applied to libraries,
generalized stencil kernels are not amenable to packaging as libraries.''

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

\begin{figure*}[!t]
\centering
\includegraphics[width=.9\linewidth]{illustrator/fig_main_R1.pdf}
\caption{
Computation speed for novel problem configurations using predictive
vs. empirical auto-tuning.
(a) As more problem configurations are used to train the regression model,
the performance of predictive auto-tuning approaches that of empirical auto-tuning
despite taking only 0.1 seconds instead of minutes.
(b) and (c) For a wide range of devices, implementation speeds found by predictive auto-tuning correlate tightly with
implementation speeds found by empirical auto-tuning.
%The boosted regression trees used for predictive auto-tuning generalize correctly across the implementation space.
}
\label{fig:fig_gflop_scatter}
\end{figure*}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{illustrator/fig_ntrain_munctional0_580.pdf} % on vader
\caption{The effect of the invalid configuration score ($\zeta$) and training set
size on predictive auto-tuning.
Models trained on as few as 25 configurations outperformed the reference.
Models trained on 50 or more configurations rivaled empirical auto-tuning.
Moderate values of $\zeta$ between $\log(0.25)$ and $\log(0.75)$ are best.
}
\label{fig:fig_ntrain}
\end{figure}

@@

While the present work serves as a basic demonstration of the value of using
machine learning models to predict optimal implementation parameters, there are
many avenues for taking these ideas further. A natural route to extend of our
current approach is to include more features as inputs to the predictive model.
In addition to further instrumentation of the kernel in question, input
features could include a much broader range of hardware-related information,
from more detailed information about device capabilities to micro-benchmarks
\citep{wong2010demystifying}, and results from performance limiter analyses
\citep{micikevicius2010analysis}. Such additional features would increase the
predictive model's ability to adapt to a wide range of different kinds of
hardware, including new devices not available at the time of kernel creation.

@@

>>>>

The paper is organized as follows: Section~\ref{sec:predautotuning}
describes the boosted regression tree model and the procedure for
fitting it to empirical timing data.  Section~\ref{sec:fbcorr} details
the data-parallel image processing algorithm we use to illustrate our
auto-tuning framework.  Section~\ref{sec:kernel} describes the sort of
kernel we employ for our benchmarking. Section~\ref{sec:exp} and
\ref{sec:results} present the results of our benchmarking experiments,
which compare a reference implementation to a) empirical auto-tuning
over a domain-specific grid, b) empirical auto-tuning over
a hill-climbing search, and c) predictive auto-tuning.
Section~\ref{sec:discuss} provides discussion and concluding remarks,
while Section~\ref{sec:fw} outlines directions for future work.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Predictive Auto-Tuning}
\label{sec:predautotuning}

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Filterbank Correlation}
\label{sec:fbcorr}

@@

The high potential for arithmetic intensity makes the GPU an ideal
platform for computing filterbank correlations, and filterbank
correlation is used extensively in image and video processing, where it
is often a computational bottleneck.  One might expect then, that it
would be easy to implement a library providing this operation as
a simple function that takes pointers and strides for $x$, $f$, and $z$
and performs the computation.  However, as shown in
\citet{pinto+cox:2011gcg} it is challenging to provide an implementation
or even an implementation strategy that provides satisfactory
performance across the range of inputs (shapes, physical layouts) that
occur in typical usage.  \citet{kamil+etal:2009} summarize a related
situation related to general stencil computations in their abstract:
``Although the auto-tuning strategy has been successfully applied to
libraries, generalized stencil kernels are not amenable to packaging as
libraries.''

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{sec:results}

\begin{figure*}[!t]
\centering
\includegraphics[width=.9\linewidth]{illustrator/fig_main_R2.pdf}
\caption{
Computation speed for novel problem configurations using predictive
vs. empirical auto-tuning.
(a) As more problem configurations are used to train the regression model,
the performance of predictive auto-tuning approaches that of empirical auto-tuning
despite taking only 0.1 seconds instead of minutes.
(b) and (c) For a wide range of devices, implementation speeds found by predictive auto-tuning correlate tightly with
implementation speeds found by empirical auto-tuning.
%The boosted regression trees used for predictive auto-tuning generalize correctly across the implementation space.
}
\label{fig:fig_gflop_scatter}
\end{figure*}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{illustrator/fig_ntrain_munctional0_580_R2.pdf} % on vader
\caption{The effect of the invalid configuration score ($\zeta$) and training set
size on predictive auto-tuning.
Models trained on as few as 25 configurations outperformed the reference.
Models trained on 50 or more configurations rivaled empirical auto-tuning.
Moderate values of $\zeta$ between $\log(0.25)$ and $\log(0.75)$ are best.
}
\label{fig:fig_ntrain}
\end{figure}

@@

While the present work serves as a basic demonstration of the value of
using machine learning models to predict optimal implementation
parameters on a specific application, there are many avenues for taking
these ideas further.

A natural route to extend of our current approach is to include more
features as inputs to the predictive model.  In addition to further
instrumentation of the kernel in question, input features could include
a much broader range of hardware-related information, from more detailed
information about device capabilities to micro-benchmarks
\citep{wong2010demystifying}, and results from performance limiter
analyses \citep{micikevicius2010analysis}. Such additional features
would increase the predictive model's ability to adapt to a wide range
of different kinds of hardware, including new devices not available at
the time of kernel creation.

@@

Other directions will include the validation of our predictive
auto-tuning framework to other domains and environments, in particular
general data-parallel primitives and non data-parallel applications,
other multi-core heterogeneous architectures, and software platforms
such as OpenCL.
